---
title: "Part 3: Quantification"
format: html
editor: visual
---

## FeatureCounts

```{r}
suppressPackageStartupMessages({
  library(tidyverse)
  library(Rsubread)
  library(stringr)
})
source("../helpers.R")
```

#### Set-up

First, we need to subset the annotation (gtf) file to include only protein-coding genes. There are a few reasons why we might rather exclude transcript types such as lncRNA and snRNA.

-   Including these other transcript types increase the number of tests, meaning more false positives, and less power to detect protein-coding genes

-   non-coding RNAs do not have poly-A tails and can't be detected with mRNA sequencing, so their absence might be misinterpreted

-   They are sometimes not well annotated

In general, a good strategy is to exclude them by default and include them by demand (i.e. if the research question explicitly demands it). Let's do this subsetting now.

```{r}
seqAnno <- getFeatureAnnotation("../../data/supplementary-files/Ensembl_R64_genes/genes_annotation_byTranscript.txt", dataFeatureType="transcript")
transcriptsUse = rownames(seqAnno)[seqAnno$type %in% "protein_coding"]
gtf <- readGff("../../data/supplementary-files/Ensembl_R64_genes/genes.gtf")
transcripts <- getGffAttributeField(gtf$attributes,
                                    field="transcript_id", 
                                    attrsep="; *", valuesep=" ")
transcriptsUse = union(transcriptsUse, setdiff(transcripts, rownames(seqAnno))) ## add those where we have no info, e.g. spikes
gtf = gtf[transcripts %in% transcriptsUse, ]

tmpOutDir <- "../../data/tmp"
# We create the tmp directory if it does not exist
dir.create(tmpOutDir, showWarnings=FALSE, recursive=TRUE)

# Save the subset GTF
gtfFile <- file.path(tmpOutDir, "genes_protein_coding.gtf")
write.table(gtf, gtfFile, quote=FALSE, sep="\t", row.names=FALSE, col.names=FALSE)
```

#### Run `featureCounts`

We are now ready to run `featureCounts` from the Rsubread package.

```{r}
bamFiles <- Sys.glob(file.path("../../results/2_mapping/STAR_alignment", "*bam"))
outputDir <- "../../results/3_quant/featureCounts"
dir.create(outputDir, showWarnings=FALSE, recursive=TRUE)

for (bamFile in bamFiles) {
  sampleName <- tools::file_path_sans_ext(basename(bamFile))
  outputPrefix <- file.path(outputDir, sampleName)
  
  countResult = Rsubread::featureCounts(
    bamFile, 
    annot.inbuilt=NULL,
    annot.ext=gtfFile,
    isGTFAnnotationFile=TRUE,
    GTF.featureType="exon",
    GTF.attrType="gene_id",
    useMetaFeatures=TRUE,
    allowMultiOverlap=TRUE,
    isPairedEnd=FALSE, 
    nthreads=4, 
    strandSpecific=2,
    minMQS=10,
    minOverlap=10,
    primaryOnly=TRUE
  )
  colnames(countResult$counts) = "matchCounts"
  data.table::fwrite(countResult$counts, file=paste0(outputPrefix, ".txt"), sep="\t", row.names=TRUE)
  colnames(countResult$stat) = c("Status", "Count")
  data.table::fwrite(countResult$stat, file=paste0(outputPrefix, "-stats.txt"), row.names=FALSE)
}
```
