---
title: "Part 1: Quality Control"
format:
  html:
    embed-resources: true
editor: visual
---

```{r}
suppressPackageStartupMessages({
  library(tidyverse)
})
```

# Introduction

Welcome to our introductory RNA-seq course.

In this course, we will be primarily working with yeast samples, initially sequenced and analyzed at the Functional Genomics Center Zürich. We will be exploring this dataset thoroughly, starting with the QC discussed in the presentations. For this first part, we will be dealing with the full data. For later parts, such as the mapping module, subsets of the data will be used instead to save computational resources and allow for faster processing.

## Downloading the data

First, let's download the data. In this directory, you will find a bash script called `01_download_data.sh`. This script will download a tar file containing our fastq files, decompress it, and deposit the files into the folder `data/tmp/yeast_fastq_full`. To execute this script, please switch to the terminal and execute the following lines (ensuring that you are in the root of the project directory, i.e. `/home/rstudio/work/rna-seq-course-feb2023`).

```{bash, eval=FALSE}
CURR_DIR=$(pwd)
cd ../..

# First we make a directory to house our data
YEAST_DIR=data/tmp/yeast_fastq_full
mkdir -p $YEAST_DIR

# Download the data
cd $YEAST_DIR
curl -O https://fgcz-gstore.uzh.ch/public/RNASeqCourse/yeast_full.tar

# Extract data
tar -xvf yeast_full.tar
rm -f yeast_full.tar

# Back home
cd $CURR_DIR
```

The data should be downloaded into `data/tmp/yeast_fastq_full/`. To check this, run

```{bash, eval=FALSE}
ls -lh ../../data/tmp/yeast_fastq_full
```

You should see 9 files in this directory. 8 files ending in fastq.gz and one "dataset.tsv" file. The "dataset.tsv" file gives some details about the samples. Let's load this file now in R to see how it looks like.

```{r}
ds <- data.table::fread("../../data/tmp/yeast_fastq_full/dataset.tsv") %>% as_tibble()
ds
```

1\. How many groups of samples are there? 2. What Library Prep Kit was used? 3. What is the read count? Is this what we expect for RNA-seq?

# FastQC

```{sh}
pwd
```
