---
title: "Part 1: Quality Control"
format:
  html:
    embed-resources: true
editor: visual
bibliography: references.bib
---

```{r}
suppressPackageStartupMessages({
  library(tidyverse)
})
```

# Introduction

Welcome to our introductory RNA-seq course.

In this part of the course, we will be primarily working with yeast samples, initially sequenced and analyzed at the Functional Genomics Center Zürich. We will be exploring this dataset thoroughly, starting with the QC discussed in the presentations. For this first part, we will be dealing with the full data. For later parts, such as the mapping module, subsets of the data will be used instead to save computational resources and allow for faster processing.

## Downloading the data

First, let's download the data. In this directory, you will find a bash script called `01_download_data.sh`. This script will download a tar file containing our fastq files, decompress it, and deposit the files into the folder `data/tmp/yeast_fastq_full`. To execute this script, please switch to the terminal and execute the following lines (ensuring that you are in the root of the project directory, i.e. `/home/rstudio/work/rna-seq-course-feb2023`).

```{bash, eval=FALSE}
CURR_DIR=$(pwd)
cd ../..

# First we make a directory to house our data
YEAST_DIR=data/tmp/yeast_fastq_full
mkdir -p $YEAST_DIR

# Download the data
cd $YEAST_DIR
curl -O https://fgcz-gstore.uzh.ch/public/RNASeqCourse/yeast_full.tar

# Extract data
tar -xvf yeast_full.tar
rm -f yeast_full.tar

# Back home
cd $CURR_DIR
```

The data should be downloaded into `data/tmp/yeast_fastq_full/`. To check this, run

```{bash, eval=FALSE}
ls -lh ../../data/tmp/yeast_fastq_full
```

You should see 9 files in this directory. 8 files ending in fastq.gz and one "dataset.tsv" file. The "dataset.tsv" file gives some details about the samples. Let's load this file now in R to see how it looks like.

```{r}
ds <- data.table::fread("../../data/tmp/yeast_fastq_full/dataset.tsv") %>% as_tibble()
ds
```

## Exercise #1

1.  How many groups of samples are there?

2.  What Library Prep Kit was used?

3.  What is the read count? Is this what we expect for RNA-seq?

4.  Is there information about the strandedness of the libraries?

::: {.callout-note appearance="simple" collapse="true"}
### Solutions

1.  2 groups: Glucose and GlycEth (i.e. Glucose+Ethanol)

2.  Lexogen mRNA Stranded Kit

3.  The samples vary between 5M to just under 20M reads. In general, we seek to get around 20M samples for bulk RNA-seq of human and mouse samples. Given that these are yeast samples and the corresponding genome is significantly smaller, even 5M reads is enough to get good results (see Count QC in chapter 3).

4.  The libraries are anti-sense.
:::

# FastQC & MultiQC

First, let's run FastQC. This tool was developed by Babraham Bioinformatics \[@Andrews:2010tn\] and is what we will primarily use for QC of our fastq files. In addition, we will run MultiQC [@ewels2016] which aims to aggregate outputs from tools like FastQC into a single report for easier viewing. Let's run these tools now.

```{bash}
CURR_DIR=$(pwd)
cd ../..

# Make directories
PROCESSED_FASTQC=results/1_qc/fastqc
PROCESSED_MULTIQC=results/1_qc/multiqc
mkdir -p $PROCESSED_FASTQC $PROCESSED_MULTIQC

# Process with FastQC
/opt/Fastqc/fastqc --extract -o $PROCESSED_FASTQC \
  -t 4 -a data/supplementary-files/adapter_list.txt --kmers 7 \
  data/tmp/yeast_fastq_full/*fastq.gz
  
# Combine result with multiqc
multiqc --outdir $PROCESSED_MULTIQC $PROCESSED_FASTQC

# Back home
cd $CURR_DIR
```
